Numerical Stability of Detonations in White Dwarf Simulations

### 2018-11-22 papers/ignition/paper.tex
Cleanup

Thermonuclear detonations are common to all current likely models of Type Ia
supernovae (SNe Ia), but how they are actually generated in progenitor systems
is still an open question. Different models predict different locations for
the detonation (e.g. in the center of a star or on the edge) and different
mechanisms for initiating the event (e.g. a prompt detonation in a white dwarf (WD)
collision versus a deflagration to detonation transition in the single degenerate
model). Common to all of the cases is a severe lack of numerical resolution in
the location where the detonation is expected to occur. The length and time scale
at which a detonation forms is orders of magnitude smaller than the resolution that
typical hydrodynamic simulations can achieve. Furthemore, the approximate equations
we use at the resolution that we can achieve are still notoriously difficult to
solve accurately. The mere presence of a detonation (or lack thereof) in a simulation
is therefore only weak evidence regarding whether a detonation would truly occur.

In this study we examine the challenges associated with simulating nuclear detonations.
The inspiration for this work comes from the literature on head-on collisions of WDs,
which can occur, for example, in certain triple star systems \citep{thompson:2011,hamers:2013}.
This problem is useful for understanding the problem of numerical simulation of nuclear
burning. Head-on WD collisions rapidly convert a significant amount of kinetic energy
into thermal energy in a small region and thus set up conditions ripe for a thermonuclear
detonation.

Early studies on WD collisions \citep{rosswog:2009,raskin:2010,loren-aguilar:2010,
hawley:2012,garcia-senz:2013} typically had effective spatial resolutions in the
range 100--500 km, and observed detonations that convert a large amount of
carbon/oxygen material into iron-group elements. These studies varied in methodology
(Lagrangian versus Eulerian evolution, nuclear network used) and did not closely agree
on the final result of the event (see Table 4 of \cite{garcia-senz:2013} for a summary).
\cite{kushnir:2013} argued that many of these simulations featured numerically unstable
evolution, ultimately caused by the zone size being significantly larger than the length
scale over which detonations form. The detonation length scale can vary widely based
on physical conditions \citep{seitenzahl:2009,garg:2017} but is generally much
smaller than 100 km. In this paper, we seek to find what numerical methods are required
to achieve converged detonations.

\section{Coupling Burning and Hydrodynamics}
\label{sec:hydrocoupling}

The numerical simulations in this paper are performed using the reacting hydrodynamics AMR
code \castro\ \citep{castro}. \cite{wdmergerI} described updates to the core hydrodynamic
algorithm since the code's original release. The goal of this section is to describe several
limitations we found when coupling the hydrodynamics module to the reactions module for detonation
problems, and the changes we made to the \castro\ code to deal with those limitations (though
most of the ideas are applicable to any adaptive mesh-based code). The challenge of accurately
capturing this coupling explains most of the difficulty in obtaining converged detonation
simulations.

\subsection{Timestep Size}
\label{sec:timestep}

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
Consider the basic outline of a single-level advance in an advection-reaction system:
\begin{enumerate}
  \item Evaluate timestep $\Delta t$ for the current advance
  \item Advance the nuclear burning network by $\Delta t / 2$
  \item Advance the hydrodynamics by $\Delta t$
  \item Advance the nuclear burning network by $\Delta t / 2$
  \item Return to Step 1
\end{enumerate}
The principal drawback of this approach is that the reactions and the hydrodynamics
can become decoupled from each other. For example, consider the example of a head-on
collision of white dwarfs. Initial nuclear burning will occur at the contact point
between the two stars. Because of the staggered updates from splitting, the evolution
effectively progresses as a cycle between burning for $\Delta t$ and getting fresh
material advected into the contact point by the hydro update for $\Delta t$.
When the collision begins, $\Delta t$ is controlled by the hydrodynamic stability
criterion, and may be large enough that it is possible for the burning advance in
Step 4 to completely burn the freshly advected material all the way to NSE.
Consequently the evolution is no longer a good approximation to smooth burning of
the in-falling material but rather separate discrete burning and hydro steps, and
the nature of the burning evolution will be quite different. This can have a
substantial effect on nucleosynthetic yields predicted by white dwarf simulations.

%TODO: add a ``Nonaka plot''

The size of the splitting error is a function of the advance timestep, and it
decreases as the timestep becomes smaller. When we add AMR in the regions of
nuclear burning, the hydrodynamic timestep (as set by the CFL criterion)
automatically becomes smaller due to the increased resolution, which mitigates
the splitting error. We have found that when we have sufficient refinement to
achieve a converged simulation, the size of the hydrodynamic timestep is already
small enough that the splitting error is minimal. For the test problem that we
will describe later, the operator splitting has no observable effect on the
physical effect we care about --- the time and location of the detonation --- at
any resolution. Consequently we do not apply any special timestep limiter based
on the rate of burning in this paper.

However, for completeness, we describe our implementation of a common approach
presented in the literature for timestep limiting based on burning, which should
be used when performing low resolution SN Ia simulations \citep{raskin:2010,hawley:2012}.
Defining the nuclear energy injection timescale $\tau_e$, and the species evolution
timescale $\tau_{X_k}$,
\begin{align}
  \tau_e &\equiv \frac{e}{|\dot{e}|} \\
  \tau_{X_k} &\equiv \frac{X_k}{|\dot{X_k}|},
\end{align}
where $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning, and $\dot{X_k}$ is an estimate of the time rate of change 
of the mass fraction of the species with index $k$, we define burning-limited 
timesteps $\Delta t_{be}$ and $\Delta t_{bX_k}$:
\begin{align}
  \Delta t_{be} &= f_{be}\, \tau_e \label{eq:timestep_e}\\
  \Delta t_{bX_k} &= f_{bX}\, \tau_{X_k}. \label{eq:timestep_X}
\end{align}
Given an estimate for $\dot{e}$, the factor $f_{be}$ determines by what 
fraction we would like to allow the internal energy to change
in the current timestep, under the assumption that $\dot{e}$ does not change from
timestep to timestep. Similarly, given an estimate for $\dot{X_k}$, the factor $f_{bX}$ 
determines the maximum change in the mass fraction of any species. To prevent species
whose abundances are small in absolute terms from controlling the timestep, we apply
the limiter based on $f_{bX}$ only to species with abundances in a given zone greater
than $10^{-3}$ (this abundance threshold can be changed by the user of the code). By making 
$f_{be}$ and $f_{bX}$ smaller, we can control the magnitude of the decoupling 
between the reactions and the hydro. A typical choice for $f_{be}$ in the
literature is in the range of 0.1 to 0.3, while to our knowledge a limiter based on
$f_{bX}$ has not been used by others performing these types of calculations.
This limiter is complementary to the limiter based on $f_{be}$, and can catch
certain burning regimes that limiting only based on changes in energy would miss.

At the start of each advance, we limit the size of the timestep to be the smaller
of the minimum hydro timestep (limited by the CFL condition), and the minimum of all the
burning timesteps across all zones. To do this, we need a method for determining 
$\dot{e}$ and $\dot{X_k}$. We choose to do this with a direct evaluation of the
right-hand-side of the burning network, given the current state at the time when
the timestep is evaluated. Other work in the literature has typically evaluated
this as a finite difference using the change over the last timestep. Our method
is more accurate in principle, but in practice we find no meaningful difference
between the limiting we propose and other methods of estimating the time derivatives.

\subsection{Timestep Retries}
\label{sec:retries}

There are many circumstances during a timestep of a hydro code where the advance
fails. As one common example, in \castro\ timesteps can be subcycled on fine levels,
so that if the refinement factor between levels is 4, we take four timesteps on
the fine level for every timestep on the coarse level, each with a quarter the
size of the coarse level timestep. Then the coarse and fine levels can be
synchronized \citep{bergercolella:1989}, which requires that we set the fine level
timestep at the beginning of the first fine level advance and have to use the same
timestep for each of the four advances. But it can occur that the velocity on the
fine level grows during these timesteps, such that during the second or later of
the four advances, the CFL criterion is no longer satisfied.

Another example of a failure mode comes from the timestep limiting described in
\autoref{sec:timestep}. The timestep will only actually satisfy the energy criterion
$\Delta t \leq f_e \tau_e$ and species criterion $\Delta t \leq f_{X_k} \tau_{X_k}$
when the estimates for $\dot{e}$ and $\dot{X_k}$ we generate are at least as large
as the actual rate of change of energy and mass fractions over the timestep. However,
this can assumption will sometimes fail during periods of runaway burning when the
rate of change of these quantities is highly nonlinear. We may not want to neglect
the errors caused by this approximation because they may build up over an extended
period of nonlinear evolution and perhaps substantially change the final results.

A third, related example comes from the burning module itself. The severe stiffness
of the nuclear reaction ODEs means that we can sometimes run into integration failures.
The most common cause of an integration failure is inability to reach the specified
error tolerance within a specified maximum number of timesteps. These are rare in
relative terms, but typically we find at least a few occur in any sufficiently long
simulation, and we either have to deal with these or abort the simulation. In the
\microphysics\ repository we have constructed a couple of ways of handling these
failures internal to the burner: we can attempt to loosen the tolerance of the ODEs
and try again; and, if that repeatedly fails, we can use a different integration method.
However, we can be a little wiser about this and note that the failures are almost always
ultimately related to the fact that the advance timestep is so large that the ODE
integrator has to solve a very challenging problem. For a sufficiently small timestep,
the size of the change that could occur during the burn would be similarly small,
and it would be much easier for the integration to succeed.

A natural solution to all of these problems is to be able to detect these failure
modes, reject the proposed timestep, and subcycle the advance within a level with
a smaller timestep. To this end, we have implemented a timestep retry option in
\castro. If a failure is detected, the timestep is cut by a factor of 2 and we
try again. If this does not work, the timestep is recursively cut until the
advance succeeds. This approach neatly addresses several different types of
problems in a single, simple framework.

\subsection{Numerically Unstable Burning}
\label{sec:unstable_burning}

\citet{kushnir:2013} point out that an inappropriate timestep is 
not the only way for the numerical discretization to cause 
severe errors in the burning. Another possible failure mode is when
the energy injection timescale
$\tau_e$ is shorter than the sound-crossing time $\tau_s$ in a zone.
When the sound-crossing time is too long, energy is built up in
a zone faster than it can be advected away by pressure waves.
This is of course a problem inherent only to numerically discretized
systems as the underlying fluid equations are continuous.
This can lead to a numerically seeded detonation caused by the
temperature building up too quickly in the zone; the detonation
may be spurious in this case. If the following condition holds,
we can be confident that a numerically seeded detonation has not occurred:
\begin{equation}
  \tau_s \leq f_{s}\, \tau_e \label{eq:burning_limiter_2}
\end{equation}
The sound crossing time, $\tau_s$, is given by $\Delta x / c_s$, 
where $c_s$ is the sound speed and $\Delta x$ is the (minimum) 
zone width. The parameter $f_{s}$ then determines the minimum
ratio of the nuclear energy generation timescale to the 
sound-crossing time. \citet{kushnir:2013} choose $f_{s} = 0.1$ 
for their simulations. \citet{kushnir:2013} implemented this
criterion by artificially limiting the magnitude of the energy
release after a burn.

>>>>

Thermonuclear detonations are common to all current likely models of Type Ia
supernovae (SNe Ia), but how they are actually generated in progenitor systems
is still an open question. Different models predict different locations for
the detonation and different mechanisms for initiating the event. Common to all
of the cases is a severe lack of numerical resolution in the location where the
detonation is expected to occur. The length and time scale at which a detonation
forms is orders of magnitude smaller than the resolution that typical multi-dimensional
hydrodynamic simulations can achieve. The mere presence of a detonation (or lack thereof)
in a simulation is therefore only weak evidence regarding whether a detonation would truly occur.

In this study we examine the challenges associated with simulating thermonuclear detonations.
The inspiration for this work comes from the literature on head-on collisions of WDs,
which can occur, for example, in certain triple star systems \citep{thompson:2011,hamers:2013}.
WD collisions rapidly convert a significant amount of kinetic energy into thermal energy and
thus set up conditions ripe for a thermonuclear detonation. Collisions are interesting
from the simulation perspective because of both this susceptibility to detonation and
the relative ease in setting them up.

Early studies on WD collisions \citep{rosswog:2009,raskin:2010,loren-aguilar:2010,
hawley:2012,garcia-senz:2013} typically had effective spatial resolutions in the
range 100--500 km, and observed detonations that convert a large amount of
carbon/oxygen material into iron-group elements. These studies varied in methodology
(Lagrangian versus Eulerian evolution, nuclear network used) and did not closely agree
on the final result of the event (see Table 4 of \cite{garcia-senz:2013} for a summary).
\cite{kushnir:2013} argued that many of these simulations featured numerically unstable
evolution, ultimately caused by the zone size being significantly larger than the length
scale over which detonations form. The detonation length scale can vary widely based
on physical conditions \citep{seitenzahl:2009,garg:2017} but is generally much
smaller than 100 km. In this paper, we attempt to find what simulation length scale
is required to achieve converged thermonuclear ignitions.

\section{Numerically Unstable Burning}
\label{sec:unstable_burning}

\citet{kushnir:2013} observe an important possible failure mode
for reacting hydrodynamics simulations. Let us define $\tau_e = e / \dot{e}$
as the nuclear energy injection timescale, and $\tau_s = \Delta x / c_s$
as the sound-crossing time in a zone (where $\Delta x$ is the grid
resolution and $c_s$ is the speed of sound).
When the sound-crossing time is too long, energy is built up in
a zone faster than it can be advected away by pressure waves.
This is of course a problem inherent only to numerically discretized
systems as the underlying fluid equations are continuous.
This can lead to a numerically seeded detonation caused by the
temperature building up too quickly in the zone; the detonation
may be spurious in this case. If $\tau_s \ll \tau_e$, we can be
confident that a numerically seeded detonation has not occurred.
In practice, we will quantify this as:
\begin{equation}
  \tau_s \leq f_{s}\, \tau_e \label{eq:burning_limiter_2}
\end{equation}
and require that $f_{s}$ is sufficiently smaller than one.
\citet{kushnir:2013} state that $f_{s} = 0.1$ is a sufficient
criterion. \citet{kushnir:2013} enforced this criterion on
their simulations by artificially limiting the magnitude of the energy
release after a burn.
